---
title: "census, hickey v cheltenham deaths"
author: "Wells"
date: "2026-03-04"
output: html_document
---

### Load libraries
```{r}
# install.packages("htmlwidgets")
# install.packages("leaflet")
#install.packages("sf")
#install.packages('DT')
library(DT)
library(tidyverse)
library(rio)
library(janitor)
library(htmlwidgets)
library(leaflet)
library(sf)
library(googlesheets4)
library(formattable)
```

# Load death certificate data
```{r}
gs4_deauth()

death_certs <- read_sheet("https://docs.google.com/spreadsheets/d/1Qu7FSUzCAeU4gdZ9BpIvfxI5kBp5rAOgEiaQ2h0ff4k/edit?gid=0#gid=0") |> 
     clean_names() |> 
      # Select all the list columns
    mutate(across(
      c(birth_date, age, birth_year),
      ~ .x |> 
      as.character() |>              # Pull it out of the 'list' box into a string
      str_remove_all("\\*") |>       # Remove those asterisks (e.g., "56*" becomes "56")
      na_if("NA") |>                 # Convert the string "NA" to a real R NA
      na_if("na") |>                 # Handle lowercase "na"
      na_if("N/A") |>                # Handle "N/A"
      as.numeric()                   # Finally, turn it into a number
  ))|> 
  mutate(
    birth_date_clean = as_datetime(birth_date) %>% as.Date(),
    birth_year = year(birth_date_clean)
    )
```

# Deaths vs Census

## Load 1930 Census
**verify we have the full run for 1930 at Cheltenham**
```{r}

census1930 <- read_sheet("https://docs.google.com/spreadsheets/d/1PSOeq5WZXk6i4HPoSHq-y2eEnm-bTx0ET3C7Sgnkzy8/edit?usp=sharing") |> 
     clean_names() |> 
  rename(birth_year = birth_like_date)


```

## Join 1930 census and death certs
```{r}

joined <- death_certs |> 
  inner_join(census1930, by=c("last"="last_name", "birth_year"))

```




# Deaths total - Tiasia and Ela

# Total deaths from annual reports, Cheltenham, Hickey
```{r}
chelt_deaths <- read_sheet("https://docs.google.com/spreadsheets/d/1qy17Gwxyb-W6P2tDqaI6jUymK4n7cmnFToNIkJ5GCI4/edit?usp=sharing", sheet="chelt_deaths") |> 
     clean_names() |> 
  mutate(across(
    # Select all the list columns
    c(number_of_inmates, total_deaths, total_escapes, total_hired_out, total_sent_to_hospital),
    ~ .x |> 
      as.character() |>              # Pull it out of the 'list' box into a string
      str_remove_all("\\*") |>       # Remove those asterisks (e.g., "56*" becomes "56")
      na_if("NA") |>                 # Convert the string "NA" to a real R NA
      na_if("na") |>                 # Handle lowercase "na"
      na_if("N/A") |>                # Handle "N/A"
      as.numeric()                   # Finally, turn it into a number
  ))

hickey_deaths <- read_sheet("https://docs.google.com/spreadsheets/d/1FnXL2BiktaoEQDh-TAzgdTDuucJfaENKQ1wVZaItdww/edit?usp=sharing") |> 
     clean_names() |> 
    mutate(across(
    # Select all the list columns
    c(year_of_report,number_of_inmates, total_deaths, total_escapes, total_hired_out, total_sent_to_hospital),
    ~ .x |> 
      as.character() |>              # Pull it out of the 'list' box into a string
      str_remove_all("\\*") |>       # Remove those asterisks (e.g., "56*" becomes "56")
      na_if("NA") |>                 # Convert the string "NA" to a real R NA
      na_if("na") |>                 # Handle lowercase "na"
      na_if("N/A") |>                # Handle "N/A"
      as.numeric()                   # Finally, turn it into a number
  ))

```

```{r}
chelt_death_cert_totals <- death_certs |> 
  filter(facility =="cheltenham") |> 
  group_by(death_year) |> 
  count(, name="death_cert_total") |> 
  arrange(desc(death_year)) |> 
  left_join(chelt_deaths, by=c("death_year"="year_of_report")) |> 
  select(death_year, total_deaths,  death_cert_total) |> 
  mutate(death_certs_bigger = if_else(death_cert_total > total_deaths, "Yes", "No"))


  
```


I am assuming the chelt_deaths is a pretty complete listof deaths except for 1918-1928. Maybe the death_certs will fill in those blams.

Next steps:
--group the cheltenham death_certs by year, compare to the chelt deaths. see if there are any years where the death_certs exceed totals for chelt deaths that year.  if so, then create a new total table where the death certs data fills in the blanks in the chelt-deaths dataframe

