---
title: "Bigrams for Cheltenham Research"
author: "Rob Wells"
date: '2025-11-24'
output: html_document
---

# Jour 772 Fall 2025:

In this exercise, you will import the Article Data Extraction results and create bigrams from the comments
The source is Cheltenham Article Data Extraction (Responses), 184 entries as of Nov 24.

```{r}
#load tidyverse, tidytext, rio,, janitor libraries

library(tidyverse)
library(tidytext)
library(rio)
library(janitor)
library(googlesheets4)

```

# Load data - pull directly from spreadsheet
```{r}
#renamed important details to details for simplicity

library(googlesheets4)

gs4_deauth()  
news <- read_sheet("https://docs.google.com/spreadsheets/d/1XFRMzN-xK1uQcvaEIOV7wuEPH6lYrvVpvgE3w6ngKPU/edit?usp=sharing", sheet = "Form Responses 1") |> 
     clean_names() |> 
  rename(details = important_details_from_article_about_the_victim_and_the_conditions_at_cheltenham_any_direct_quotes_need_to_have_quotation_marks)

```

# Tokenize one column

```{r}

stories <- str_replace_all(news$details, "- ", "")
stories_df <- tibble(stories,)

# unnest includes lower, punct removal

stories_tokenized <- stories_df %>%
  unnest_tokens(word,stories)

stories_tokenized
#A list of the common words
```

#List of top words, cleaned with stop words removed
```{r}

data(stop_words)

stories_tokenized <- stories_tokenized %>%
  anti_join(stop_words, by = c("word" = "word")) %>%
  filter(word != "temp_file") %>%
  filter(word != "stories_corpus") %>%
  filter(!grepl('[0-9]', word))

# Word Count of top words

word_count <- stories_tokenized %>%
  count(word, sort=TRUE)


```

# Bigrams without stop words

```{r}
bigrams <- stories_df %>%
  unnest_tokens(bigram, stories, token="ngrams", n=2) |> 
  separate(bigram, c("word1", "word2"), sep = " ")

count_bigrams <- bigrams %>%
  count(word1, word2, sort = TRUE)


```

# Bigrams with stop words

```{r}
bigrams_clean <-stories_df %>%
  unnest_tokens(bigram, stories, token="ngrams", n=2) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) |> 
  filter(!is.na(word1) | !is.na(word2))

count_clean_bigrams <- bigrams_clean  %>%
  count(word1, word2, sort = TRUE)


```

# Next steps: Categorize the material
Category "origins" = how the boys arrived -- sentenced, committed etc
"labor" - working on a farm, broom factory anything involving their work
"boy offenses" = specifies crimes the boys committed to get sentenced to Cheltenham
"investigations" = grand jury or newspaper investigations of Cheltenham
"abuse" = details of abuse, whippings, pubishment, no shoes, 
"medical" = discussing medical conditions

Examine the bigrams and top words and on the categories below


```{r}
news <- news |> 
  mutate(details_clean = str_squish(str_remove_all(details, "[[:punct:]]"))) |> 
  mutate(category = paste(
    if_else(str_detect(details_clean, "(?i)sentenced"), "origins", NA_character_),
    if_else(str_detect(details_clean, "(?i)committed"), "origins", NA_character_),
    if_else(str_detect(details_clean, "(?i)labor"), "labor", NA_character_),
    if_else(str_detect(details_clean, "(?i)farm"), "labor", NA_character_),
    if_else(str_detect(details_clean, "(?i)broom"), "labor", NA_character_),
    if_else(str_detect(details_clean, "(?i)larceny"), "boy_offenses", NA_character_),
    if_else(str_detect(details_clean, "(?i)conditions"), "treatment", NA_character_),
    if_else(str_detect(details_clean, "(?i)jury"), "investigations", NA_character_),
    if_else(str_detect(details_clean, "(?i)beaten"), "abuse", NA_character_),
    sep = "; "
  )) |> 
  mutate(category = str_remove_all(category, "NA; |; NA")) |> 
  mutate(category = na_if(category, "NA"))
```



# Next steps: Team up:
Akira, Brandyn
Nicole, Bex, Raphael
Sophie, David, Zaka

1) bigrams and word count on  "elaborate_on_the_quality_of_the_journalism_sourcing_graphics_photos_interviews"; "notes_on_neglect_or_abuse"; "notes_anything_else_thats_noteworthy"     

2) add to the catgory code above -- be specific on what you are adding so I can merge it into the master code block

