---
title: "Clean_Analyze_Death_Certificates"
author: "Raphael-Wells"
date: "2025-12-4"
output: html_document
---

This is Raphael Romero Ruiz's Dec 4 Death Certificate cleaning code
The source comes from this assignment: https://umd.instructure.com/courses/1389907/assignments/7403074?module_item_id=14042327



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries
```{r}
# install.packages("htmlwidgets")
# install.packages("leaflet")
#install.packages("sf")
#install.packages('DT')
library(DT)
library(tidyverse)
library(rio)
library(janitor)
library(htmlwidgets)
library(leaflet)
library(sf)
library(googlesheets4)
library(formattable)
```

### Import Extracted Death Certificate Data
```{r}

gs4_deauth()  
dc <- read_sheet("https://docs.google.com/spreadsheets/d/13CthhhfLHiPZqr1P67SCa9BVUyb5Rl6OBhUwvqawDEk/edit?usp=sharing", sheet = "master_deaths_dec_5", col_types ="c") |> 
     clean_names() |> 
    mutate(approx_age = as.numeric(str_extract(age_on_death_certificate, "^\\d+")),
         death_date1 = death_date)


#imports 176 entries - Dec 4


```


```{r include=FALSE}
# Function to parse the different date formats
standardize_date <- function(date_string) {
  if (is.na(date_string) || date_string == "N/A") {
    return(NA)
  }
  
  # Try parsing as "d Mon yyyy" format first
  parsed <- dmy(date_string, quiet = TRUE)
  
  # If that fails, try "m/d/yyyy" format
  if (is.na(parsed)) {
    parsed <- mdy(date_string, quiet = TRUE)
  }
  
  # If still fails, try "Month yyyy" format (will default to 1st of month)
  if (is.na(parsed)) {
    parsed <- my(date_string, quiet = TRUE)
  }
  
  return(parsed)
}

dc$death_date_clean <- sapply(dc$death_date1, standardize_date)
dc$death_date_clean <- as.Date(dc$death_date_clean)
```



### Death year
```{r}
dc <- dc |> 
  mutate(death_year = year(death_date_clean),
         death_decade = floor(death_year / 10) * 10)
```

#### Birth dates
Using the date cleaning sequence above, do the same processing for birth dates**

```{r}
standardize_birth_date <- function(date_string) {
  if (is.na(date_string) || date_string == "N/A") {
    return(NA)
  }
  
  # Try parsing as "d Mon yyyy" format first
  parsed <- dmy(date_string, quiet = TRUE)
  
  # If that fails, try "m/d/yyyy" format
  if (is.na(parsed)) {
    parsed <- mdy(date_string, quiet = TRUE)
  }
  
  # If still fails, try "Month yyyy" format (will default to 1st of month)
  if (is.na(parsed)) {
    parsed <- my(date_string, quiet = TRUE)
  }
  
  # Fix 2-digit year issues - if year > 1950, assume it's 1800s
  if (!is.na(parsed)) {
    year_val <- year(parsed)
    if (year_val > 1950) {
      parsed <- parsed - years(100)
    }
  }
  
  return(parsed)
}

dc$birth_date_clean <- sapply(dc$birth_date, standardize_birth_date)
dc$birth_date_clean <- as.Date(dc$birth_date_clean)

dc <- dc |> 
  mutate(birth_year = year(birth_date_clean))
```


## Check for duplicate names
```{r}

dc |> 
  count(name) |> 
  filter(n > 1)

#James Evans. One entry #39 and the other #199. Both seem separate
```

mark duplicates
```{r}

# dupes <- dc |> 
#   count(name) |> 
#   filter(n > 1) |> 
#   select(name) |> 
#   mutate(name = str_squish(str_replace_all(name,'\\.',''))) |> 
#   pull(name)

# 
# dc <- dc |> 
#   mutate(dupes = case_when(
#     name %in% dupes ~ "x")
#   ) |> 
#   arrange(personal_id) |> 
#   select(dupes, everything()) 


```

# Primary Cause of Death


```{r}
# Combine cause of death columns, cleaning up semicolons and "NA" entries
dc <- dc |> 
   mutate(primary_death_update = primary_death) |> 
   mutate(primary_death_update= str_replace_all(primary_death_update, "(?i) & ", "; ")) |>
  mutate(primary_death_update= str_replace_all(primary_death_update, "NA", "")) |> 
  #removing pesky leading, trailing semicolons
  mutate(primary_death_update= str_remove_all(primary_death_update, "^;+|;+$")) |> 
  mutate(primary_death_update= str_remove_all(primary_death_update, "^[;\\s]+|[;\\s]+$")) |> 
  mutate(primary_death_update= str_replace_all (primary_death_update, "blank", "")) |> 
  mutate(primary_death_update= na_if(primary_death_update, "")) 

```

# Primary causes of death

```{r}
all_causes <- dc %>%
  select(primary_death_update) %>%
  filter(!is.na(primary_death_update)) %>%
  distinct() %>%
  arrange(primary_death_update)
#87 primary causes
#for all 114 different causes and groupings Dec 5
```

# Primary categories

```{r}

dc <- dc %>% 
  mutate(
    tuberculosis = str_detect(primary_death_update, "(?i)tubercul|phthisis|pthisis|consumption|terberculosis"),
    pneumonia = str_detect(primary_death_update, "(?i)pneumonia"), 
    typhoid = str_detect(primary_death_update, "(?i)typhoid"),
    asthma = str_detect(primary_death_update, "(?i)asthma"),
    meningitis = str_detect(primary_death_update, "(?i)meningitis"),
    peritonitis = str_detect(primary_death_update, "(?i)peritonit"),
    cardiac_issues = str_detect(primary_death_update, "(?i)cardiac|heart failure|heart trouble|mitral"),
    gastric_ulcer = str_detect(primary_death_update, "(?i)gastric.*ulcer|ulcer.*gastric|gastroenteritis"),
    encephalitis = str_detect(primary_death_update, "(?i)encephalitis"),
    malaria = str_detect(primary_death_update, "(?i)malaria"),
    asphyxia = str_detect(primary_death_update, "(?i)asphyxia"),
    exhaustion = str_detect(primary_death_update, "(?i)exhaust"),  # Catches both exhaustion and exhaution typo
    asthenia = str_detect(primary_death_update, "(?i)asthenia"),
    lung_disease = str_detect(primary_death_update, "(?i)lungs|oedema of lungs"),
    hemorrhage = str_detect(primary_death_update, "(?i)hemo?rrhage|hemm(?!ophilia)"),  # Fixed to exclude hemophilia
    influenza = str_detect(primary_death_update, "(?i)influenza"),
    nephritis = str_detect(primary_death_update, "(?i)nephritis"),
    septicemia = str_detect(primary_death_update, "(?i)septic"),
    diptheria = str_detect(primary_death_update, "(?i)diptheria"),
    drowning = str_detect(primary_death_update, "(?i)drowning"),
    appendicitis = str_detect(primary_death_update, "(?i)appendicitis"),
    spine_disease = str_detect(primary_death_update, "(?i)disease of the spine|spinal"),
    blood_disease = str_detect(primary_death_update, "(?i)leusocythacenia|leuko|hemophilia"),  # NEW - for leukemia variations and hemophilia
    violent_injury = str_detect(primary_death_update, "(?i)wound.*groin|brick|Brain|bullet|shot|stab*|wound.*heart"), 
     premature_birth = str_detect(primary_death_update,"(?i)Prematurely 6 weeks gestation|Immaturely \\(6 month baby\\)"), # NEW - premature birth
    no_details = is.na(primary_death_update) | str_detect(primary_death_update, "(?i)\\bNA\\b|N/A|missing|illegible|information|no_details|unknown")
  )

```


### Calculate death causes by year

```{r}
# Summarize causes of death by year in long format - UPDATED with new categories
summary_long <- dc |> 
  group_by(death_year) |> 
  summarise(
    tuberculosis = sum(tuberculosis, na.rm = TRUE),
    pneumonia = sum(pneumonia, na.rm = TRUE),
    typhoid = sum(typhoid, na.rm = TRUE),
    asthma = sum(asthma, na.rm = TRUE),
    meningitis = sum(meningitis, na.rm = TRUE),
    peritonitis = sum(peritonitis, na.rm = TRUE),
    cardiac_issues = sum(cardiac_issues, na.rm = TRUE),
    exhaustion = sum(exhaustion, na.rm = TRUE),
    gastric_ulcer = sum(gastric_ulcer, na.rm = TRUE),
    encephalitis = sum(encephalitis, na.rm = TRUE),
    malaria = sum(malaria, na.rm = TRUE),
    asphyxia = sum(asphyxia, na.rm = TRUE),
    asthenia = sum(asthenia, na.rm = TRUE),
    lung_disease = sum(lung_disease, na.rm = TRUE),
    hemorrhage = sum(hemorrhage, na.rm = TRUE),
    influenza = sum(influenza, na.rm = TRUE),
    nephritis = sum(nephritis, na.rm = TRUE),
    septicemia = sum(septicemia, na.rm = TRUE),
    diptheria = sum(diptheria, na.rm = TRUE),
    drowning = sum(drowning, na.rm = TRUE),
    appendicitis = sum(appendicitis, na.rm = TRUE),
    spine_disease = sum(spine_disease, na.rm = TRUE),
    blood_disease = sum(blood_disease, na.rm = TRUE),
    premature_birth = sum(premature_birth, na.rm = TRUE),
    violent_injury = sum(violent_injury, na.rm = TRUE),
    no_details = sum(no_details, na.rm = TRUE)
  ) |> 
  pivot_longer(cols = -death_year, 
               names_to = "cause", 
               values_to = "count")
```

# Death Categories summary
```{r}
death_categories_summary <- summary_long |> 
  group_by(cause) |> 
  summarise(total = sum(count)) |>
  arrange(desc(total))

#write_csv(death_categories_summary, "output/primary_death_categories_summary_dec5.csv")

summary_long |> 
  group_by(cause) |> 
  summarise(total = sum(count)) |>
  arrange(desc(total)) |>
  datatable(caption = "Primary Causes of Death, Cheltenham, Jessup",
            options = list(pageLength = 50))
```


# Primary cause Higher level categories
```{r}
# Take the Higher Level Categories and create 9 new columns, one per category 

dc <- dc %>% 
  mutate(
    cat_respiratory = tuberculosis | pneumonia | asthma | influenza | asphyxia | lung_disease | 
                     str_detect(primary_death_update, "(?i)Suffocation"),
    cat_infection = meningitis | peritonitis | encephalitis | malaria | septicemia | diptheria,
    cat_intestinal_kidney = typhoid | gastric_ulcer | appendicitis | nephritis | str_detect(primary_death_update, "(?i)Chronic Nephritis|acute parenchymatous nephritis anasarca | stomach |Bright's disease"),
    cat_cardiac = cardiac_issues,
    cat_violent = violent_injury,
    cat_trauma = hemorrhage | drowning,
    cat_premature_birth = premature_birth,
    cat_exhaustion = exhaustion | asthenia,
    cat_other =  blood_disease  | spine_disease |               str_detect(primary_death_update, "(?i)No further information given | Spinal Paralysis"),
    cat_no_details = no_details
  )
```


## Check higher level porimary categories totals
```{r}
# Summing higher level categories totals
primary_summary <- dc %>%
  summarise(
    cat_respiratory = sum(cat_respiratory, na.rm = TRUE),
    cat_infection = sum(cat_infection, na.rm = TRUE),
    cat_intestinal_kidney = sum(cat_intestinal_kidney, na.rm = TRUE),
    cat_cardiac = sum(cat_cardiac, na.rm = TRUE),
    cat_violent = sum(cat_violent, na.rm = TRUE),
    cat_trauma = sum(cat_trauma, na.rm = TRUE),
    cat_exhaustion = sum(cat_exhaustion, na.rm = TRUE),
    cat_other = sum(cat_other, na.rm = TRUE),
    cat_premature_birth = sum(cat_premature_birth, na.rm = TRUE),
    cat_no_details = sum(cat_no_details, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "category", values_to = "count") %>%
  arrange(desc(count))
primary_summary 
```


### Table: Primary Cause of Death Summary
```{r}
primary_summary |> 
  mutate(category = str_to_sentence(str_remove_all(category, "cat_"))) |> 
  rename(Primary_Cause_Death = category, Total = count) |> 
  mutate(Primary_Cause_Death = fct_reorder(Primary_Cause_Death, Total)) |> 
  ggplot(aes(x = Total, y = Primary_Cause_Death, fill = Total)) +
  geom_col(position = "dodge") +
  theme(legend.position = "none") +
  geom_text(aes(label = (Total)), size = 4, hjust =-.1) +
  scale_x_continuous(breaks = seq(0, 130, 10), limits=c(0, 130)) +
  labs(title = "Summary Primary Causes of Death at Cheltenham, Jessup",
       subtitle = "From 1880-1941",
       caption = "Graphic by Rob Wells, 12-5-2025",
       y = "",
       x = "")
  

```






#----------------------------------------------------#

### **ALL** Causes of death

```{r}
# Combine cause of death columns, cleaning up semicolons and "NA" entries
dc2 <- dc |> 
  # mutate(cause_death_update = paste(cause_of_death, if_cause_of_death_different_put_that_here, sep = "; ")) |> 
   mutate(cause_death_update = cause_of_death) |> 
   mutate(cause_death_update = str_replace_all(cause_death_update, "(?i) & ", "; ")) |>
  mutate(cause_death_update = str_replace_all(cause_death_update, "NA", "")) |> 
  #removing pesky leading, trailing semicolons
  mutate(cause_death_update = str_remove_all(cause_death_update, "^;+|;+$")) |> 
  mutate(cause_death_update = str_remove_all(cause_death_update, "^[;\\s]+|[;\\s]+$")) |> 
  mutate(cause_death_update = str_replace_all (cause_death_update, "blank", "")) |> 
  mutate(cause_death_update = na_if(cause_death_update, "")) 

```


# Categorize causes of death

```{r}
all_causes_dc2 <- dc2 %>%
  select(cause_death_update) %>%
  filter(!is.na(cause_death_update)) %>%
  distinct() %>%
  arrange(cause_death_update)
#114 different causes and groupings Dec 5
```

```{r}

dc2 <- dc2 %>% 
  mutate(
    tuberculosis = str_detect(cause_death_update, "(?i)tubercul|phthisis|pthisis|consumption|terberculosis"),
    pneumonia = str_detect(cause_death_update, "(?i)pneumonia"), 
    typhoid = str_detect(cause_death_update, "(?i)typhoid"),
    asthma = str_detect(cause_death_update, "(?i)asthma"),
    meningitis = str_detect(cause_death_update, "(?i)meningitis"),
    peritonitis = str_detect(cause_death_update, "(?i)peritonit"),
    cardiac_issues = str_detect(cause_death_update, "(?i)cardiac|heart failure|heart trouble|mitral"),
    gastric_ulcer = str_detect(cause_death_update, "(?i)gastric.*ulcer|ulcer.*gastric|gastroenteritis"),
    encephalitis = str_detect(cause_death_update, "(?i)encephalitis"),
    malaria = str_detect(cause_death_update, "(?i)malaria"),
    asphyxia = str_detect(cause_death_update, "(?i)asphyxia"),
    exhaustion = str_detect(cause_death_update, "(?i)exhaust"),  # Catches both exhaustion and exhaution typo
    asthenia = str_detect(cause_death_update, "(?i)asthenia"),
    lung_disease = str_detect(cause_death_update, "(?i)lungs|oedema of lungs"),
    hemorrhage = str_detect(cause_death_update, "(?i)hemo?rrhage|hemm(?!ophilia)"),  # Fixed to exclude hemophilia
    influenza = str_detect(cause_death_update, "(?i)influenza"),
    nephritis = str_detect(cause_death_update, "(?i)nephritis"),
    septicemia = str_detect(cause_death_update, "(?i)septic"),
    diptheria = str_detect(cause_death_update, "(?i)diptheria"),
    drowning = str_detect(cause_death_update, "(?i)drowning"),
    appendicitis = str_detect(cause_death_update, "(?i)appendicitis"),
    spine_disease = str_detect(cause_death_update, "(?i)disease of the spine|spinal"),
    blood_disease = str_detect(cause_death_update, "(?i)leusocythacenia|leuko|hemophilia"),  # NEW - for leukemia variations and hemophilia
    violent_injury = str_detect(cause_death_update, "(?i)wound.*groin|brick|Brain|bullet|shot|stab*|wound.*heart"), 
     premature_birth = str_detect(cause_death_update,"(?i)Prematurely 6 weeks gestation|Immaturely \\(6 month baby\\)"), # NEW - premature birth
    no_details = is.na(cause_death_update) | str_detect(cause_death_update, "(?i)\\bNA\\b|N/A|missing|illegible|information|no_details|unknown")
  )
```

### Calculate death causes by year

```{r}
# Summarize causes of death by year in long format - UPDATED with new categories
summary_long_dc2 <- dc2 |> 
  group_by(death_year) |> 
  summarise(
    tuberculosis = sum(tuberculosis, na.rm = TRUE),
    pneumonia = sum(pneumonia, na.rm = TRUE),
    typhoid = sum(typhoid, na.rm = TRUE),
    asthma = sum(asthma, na.rm = TRUE),
    meningitis = sum(meningitis, na.rm = TRUE),
    peritonitis = sum(peritonitis, na.rm = TRUE),
    cardiac_issues = sum(cardiac_issues, na.rm = TRUE),
    exhaustion = sum(exhaustion, na.rm = TRUE),
    gastric_ulcer = sum(gastric_ulcer, na.rm = TRUE),
    encephalitis = sum(encephalitis, na.rm = TRUE),
    malaria = sum(malaria, na.rm = TRUE),
    asphyxia = sum(asphyxia, na.rm = TRUE),
    asthenia = sum(asthenia, na.rm = TRUE),
    lung_disease = sum(lung_disease, na.rm = TRUE),
    hemorrhage = sum(hemorrhage, na.rm = TRUE),
    influenza = sum(influenza, na.rm = TRUE),
    nephritis = sum(nephritis, na.rm = TRUE),
    septicemia = sum(septicemia, na.rm = TRUE),
    diptheria = sum(diptheria, na.rm = TRUE),
    drowning = sum(drowning, na.rm = TRUE),
    appendicitis = sum(appendicitis, na.rm = TRUE),
    spine_disease = sum(spine_disease, na.rm = TRUE),
    blood_disease = sum(blood_disease, na.rm = TRUE),
    premature_birth = sum(premature_birth, na.rm = TRUE),
    violent_injury = sum(violent_injury, na.rm = TRUE),
    no_details = sum(no_details, na.rm = TRUE)
  ) |> 
  pivot_longer(cols = -death_year, 
               names_to = "cause", 
               values_to = "count")
```

# Death Categories summary
```{r}
dc2_summary <- summary_long_dc2 |> 
  group_by(cause) |> 
  summarise(total = sum(count)) |>
  arrange(desc(total))

summary_long_dc2 |> 
  group_by(cause) |> 
  summarise(total = sum(count)) |>
  arrange(desc(total)) |>
  datatable(caption = "All Causes of Death, Cheltenham, Jessup",
            options = list(pageLength = 50))
```


# Higher level categories
```{r}
# Take the Higher Level Categories and create 9 new columns, one per category 

dc2 <- dc2 %>% 
  mutate(
    cat_respiratory = tuberculosis | pneumonia | asthma | influenza | asphyxia | lung_disease | 
                     str_detect(cause_death_update, "(?i)Suffocation"),
    cat_infection = meningitis | peritonitis | encephalitis | malaria | septicemia | diptheria,
    cat_intestinal_kidney = typhoid | gastric_ulcer | appendicitis | nephritis | str_detect(cause_death_update, "(?i)Chronic Nephritis|acute parenchymatous nephritis anasarca | stomach |Bright's disease"),
    cat_cardiac = cardiac_issues,
    cat_violent = violent_injury,
    cat_trauma = hemorrhage | drowning,
    cat_premature_birth = premature_birth,
    cat_exhaustion = exhaustion | asthenia,
    cat_other =  blood_disease  | spine_disease |               str_detect(cause_death_update, "(?i)No further information given | Spinal Paralysis"),
    cat_no_details = no_details
  )
```

# Data quality checks

## Check higher level categories totals
```{r}
# Summing higher level categories totals
dc2 %>%
  summarise(
    cat_respiratory = sum(cat_respiratory, na.rm = TRUE),
    cat_infection = sum(cat_infection, na.rm = TRUE),
    cat_intestinal_kidney = sum(cat_intestinal_kidney, na.rm = TRUE),
    cat_cardiac = sum(cat_cardiac, na.rm = TRUE),
    cat_violent = sum(cat_violent, na.rm = TRUE),
    cat_trauma = sum(cat_trauma, na.rm = TRUE),
    cat_exhaustion = sum(cat_exhaustion, na.rm = TRUE),
    cat_other = sum(cat_other, na.rm = TRUE),
    cat_premature_birth = sum(cat_premature_birth, na.rm = TRUE),
    cat_no_details = sum(cat_no_details, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "category", values_to = "count") %>%
  arrange(desc(count))
```


## Check for records with no category assigned
```{r}
# Find records that don't match any higher level category
no_cat <- dc2 %>%
  filter(!cat_respiratory & !cat_infection & !cat_intestinal_kidney & 
         !cat_cardiac & !cat_trauma & !cat_exhaustion & !cat_other & !cat_no_details & !cat_violent & !cat_premature_birth) %>%
  select(name, cause_death_update, tuberculosis:no_details)
```

## Compare asphyxia detection
```{r}
# Check if asphyxia is now being detected properly
dc2 %>%
  filter(asphyxia == TRUE) %>%
  select(name, cause_death_update, asphyxia, cat_respiratory)
```

## Check cardiac
```{r}

cardiac <- dc2 |> 
  filter(cardiac_issues=="TRUE") |> 
  select(facility, personal_id, name, death_date_clean, death_year, cause_of_death, cardiac_issues, age_on_death_certificate)

cardiac |> 
  group_by(death_year) |> 
  count() |> 
 ggplot(aes(x = death_year, y = n, fill = n)) +
  geom_col(position = "dodge") +
  theme(legend.position = "none") +
  labs(title = "Cardiac Related Deaths at Cheltenham, Jessup",
       subtitle = "From 1880-1941",
       caption = "Graphic by Rob Wells, 12-4-2025",
       y="Number",
       x="Year")

```


#respiratory
##tuberculosis
##pneumonia
##asthma
##influenza
##Immediate Asphyxia #(complications from phthisis)
##Spinal Paralysis #(oedema of the lungs)
##Suffocation #(pneumonia)
##lung_disease


#intenstinal
##typhoid
##gastric_ulcer


#cardiac
##cardiac_issues

#trauma
##hemorrhage
##Compression of the Brain


#infection
##meningitis
##peritonitis
##No further information given #peritonitis
##encephalitis
##Diptheria

#kidney disease
##Chronic Nephritis
##acute parenchymatous nephritis anasarca

#other
##exhaustion
##delirium
##asthenia
##Accidental Drowning
##Prematurely 6 weeks gestation
##Immaturely 6 month baby



#Graphics




### Tuberculosis by year
```{r}

summary_long |> 
  filter(cause == "tuberculosis") |> 
  group_by(death_year) |> 
 ggplot(aes(x = death_year, y = count, fill = count)) +
  geom_col(position = "dodge") +
  theme(legend.position = "none") +
  labs(title = "Tuberculosis Deaths at Cheltenham, Jessup",
       subtitle = "From 1880-1941",
       caption = "Graphic by Rob Wells, 12-4-2025",
       y="Number",
       x="")


```



# Broad category by year
```{r}
#RR was here
# Summarize by broad category and year
category_summary <- category_new %>%
  group_by(death_year, category_broad) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(category_broad))

# Create bar chart
ggplot(category_summary, aes(x = death_year, y = total_count, fill = category_broad)) +
  geom_col(position = "stack") +
  labs(title = "Causes of Death by Category at Cheltenham, Jessup",
       subtitle = "Totals include multiple entries for individuals with more than one cause of death.",
       caption = "Graphic by Raphael Romero Ruiz, 12-2-2025",
       y = "Number",
       x = "Year",
       fill = "Category") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11),
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("category_summary.png",  width = 12, height = 6, dpi = 300)
```


### 2. Infectious Disease Trends - Line Chart with Annotations
#rsw comment: make into a bar chart
```{r}
summary_long |>
  filter(cause %in% c("tuberculosis", "pneumonia", "typhoid", "meningitis", "influenza")) |>
  ggplot(aes(x = death_year, y = count, color = cause, group = cause)) +
  geom_line(linewidth = 1.2, alpha = 0.9) +
  # Add annotation for Spanish Flu
  annotate("rect", xmin = 1918, xmax = 1920, ymin = -Inf, ymax = Inf, 
           alpha = 0.15, fill = "red") +
  annotate("text", x = 1919, y = 9.5, 
           label = "Spanish Flu\nPandemic", size = 3, fontface = "italic", color = "darkred") +
  scale_color_brewer(palette = "Set1",
                     labels = c("Influenza", "Meningitis", "Pneumonia", 
                               "Tuberculosis", "Typhoid")) +
  scale_x_continuous(breaks = seq(1880, 1940, by = 10),
                     limits = c(1875, 1945),  # Add padding on both sides
                     expand = expansion(mult = 0.02)) +  # Extra spacing
  scale_y_continuous(breaks = 1:10,  # Whole numbers 1-10
                     limits = c(0, 10)) +
  labs(
    title = "Infectious Disease Deaths at Cheltenham, Jessup",
    subtitle = "Tracking major infectious diseases from 1880-1941",
    caption = "Graphic by Sophia Hernandez-Piña and Zaka Hossain",
    x = "Year",
    y = "Number of Deaths",
    color = "Disease"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    plot.caption = element_text(hjust = 0, color = "gray50"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.margin = margin(10, 20, 10, 10)  # Extra margin for spacing
  )

#ggsave("infectious_diseases.png", width = 12, height = 6, dpi = 300)
```


### 3. Comparative Heatmap - Disease Intensity Over Time
#rsw comment : we will remake make this by decade but will have to wait until monday.
```{r}
summary_long |>
  filter(cause %in% c("tuberculosis", "pneumonia", "typhoid", 
                      "meningitis", "influenza", "cardiac_issues",
                      "heart_failure", "hemorrhage")) |>
  mutate(cause_label = str_to_title(str_replace_all(cause, "_", " "))) |>
  ggplot(aes(x = death_year, y = cause_label, fill = count)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(count > 0, count, "")), 
            color = "white", size = 3, fontface = "bold") +
  scale_fill_gradient2(
    low = "#f7f7f7", 
    mid = "#fc8d59", 
    high = "#b30000",
    midpoint = 5,
    na.value = "white",
    name = "Deaths"
  ) +
  scale_x_continuous(breaks = seq(1880, 1940, by = 5), expand = c(0,0)) +
  # Add vertical line for Spanish Flu
  geom_vline(xintercept = c(1918, 1920), color = "darkred", 
             linetype = "dashed", linewidth = 0.8, alpha = 0.5) +
  labs(
    title = "Disease Mortality Heatmap: Cheltenham, Jessup (1880-1941)",
    subtitle = "Darker colors indicate higher death counts | Dashed vertical lines mark Spanish Flu pandemic (1918-1920)",
    caption = "Graphic by Sophia Hernandez-Piña and Zaka Hossain",
    x = "Year",
    y = "Cause of Death"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    plot.caption = element_text(hjust = 0, color = "gray50"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    panel.grid = element_blank(),
    legend.position = "right"
  )
#ggsave("heat_map_infectious_diseases.png", width = 12, height = 6, dpi = 300)
```

# Average Annual Deaths by Time Period
#make this a tuberculosis only graphic

```{r}
summary_long |>
  filter(cause %in% c("tuberculosis", "pneumonia", "influenza")) |>
  mutate(
    period = case_when(
      death_year < 1918 ~ "Pre-Pandemic (1880-1917)",
      death_year >= 1918 & death_year <= 1920 ~ "Pandemic (1918-1920)",
      death_year > 1920 ~ "Post-Pandemic (1921-1941)"
    ),
    # Reorder the periods
    period = factor(period, levels = c("Pre-Pandemic (1880-1917)", 
                                       "Pandemic (1918-1920)", 
                                       "Post-Pandemic (1921-1941)")),
    cause_label = case_when(
      cause == "tuberculosis" ~ "Tuberculosis",
      cause == "pneumonia" ~ "Pneumonia",
      cause == "influenza" ~ "Influenza"
    )
  ) |>
  group_by(period, cause_label) |>
  summarise(avg_deaths = mean(count, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = cause_label, y = avg_deaths, fill = period)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = round(avg_deaths, 1)), 
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_fill_manual(
    values = c("Pre-Pandemic (1880-1917)" = "skyblue",
               "Pandemic (1918-1920)" = "red",
               "Post-Pandemic (1921-1941)" = "lightblue")
  ) +
  labs(
    title = "Average Annual Deaths by Time Period",
    subtitle = "Comparing respiratory disease mortality before, during, and after the Spanish Flu pandemic",
    caption = "Graphic by Sophia Hernandez-Piña and Zaka Hossain",
    x = "Disease",
    y = "Average Deaths Per Year",
    fill = "Time Period"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    plot.caption = element_text(hjust = 0, color = "gray50"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
#ggsave("pandemic_deaths_comparison.png", width = 12, height = 6, dpi = 300)
```



## Mapping place of origin

```{r}

dc |> 
  mutate(origin = birth_place_city_or_town_and_state_write_it_as_bowie_md) |> 
  count(origin) |> 
  arrange(desc(n))


```

### clean origin

```{r}
dc <- dc |> 
  mutate(origin = str_squish(birth_place_city_or_town_and_state_write_it_as_bowie_md)) |> 
  mutate(county_origin = case_when(
         str_detect(origin, "(?i)^md\\.?$") ~ "Maryland", 
         str_detect(origin, "(?i)baltimore county") ~ "Baltimore County",  # Check this FIRST
         str_detect(origin, "(?i)baltimore") ~ "Baltimore city",
         str_detect(origin, "(?i)easton") ~ "Talbot County",
         str_detect(origin, "(?i)frederick") ~ "Frederick County",
         str_detect(origin, "(?i)Prince George's County") ~ "Prince George's County",
         str_detect(origin, "(?i)Prince Georges County, Md.") ~ "Prince George's County",
         str_detect(origin, "(?i)Seat Pleasant") ~ "Prince George's County",
         str_detect(origin, "(?i)Cheltenham") ~ "Prince George's County",
         str_detect(origin, "(?i)unknown") ~ "Unknown",
         str_detect(origin, "(?i)Centreville") ~ "Queen Anne's County",
         str_detect(origin, "(?i)church hill") ~ "Queen Anne's County",
         str_detect(origin, "(?i)Salisbury") ~ "Wicomico County",
         str_detect(origin, "(?i)Pocomoke") ~ "Worcester County",
         str_detect(origin, "(?i)Anne Arrundel County") ~ "Anne Arundel County",
         str_detect(origin, "(?i)Chaptico, St. Marys County, Md.") ~ "St. Mary's County",
         str_detect(origin, "(?i)Dunkirk, Md.") ~ "Calvert County",
         str_detect(origin, "(?i)Elkton, Md.") ~ "Cecil County",
        str_detect(origin, "(?i)Ellicott City MD") ~ "Howard County",
        str_detect(origin, "(?i)Ellicott") ~ "Howard County",
        str_detect(origin, "(?i)Harford County") ~ "Harford County",
        str_detect(origin, "(?i)Illegible") ~ "Unknown",
        str_detect(origin, "(?i)Jessup") ~ "Anne Arundel County", 
        str_detect(origin, "(?i)oakland") ~ "Garrett County",
        str_detect(origin, "(?i)Md. former residence Pocomoke, Md.") ~ "Worcester County",
        # Counties explicitly mentioned
         str_detect(origin, "(?i)prince george") ~ "Prince George's County",
         str_detect(origin, "(?i)anne arr") ~ "Anne Arundel County",  # Catches misspelling
         str_detect(origin, "(?i)harford county") ~ "Harford County",
         str_detect(origin, "(?i)chaptico") ~ "St. Mary's County",
       # str_detect(origin, "St.Marys County") ~ "St. Mary's County",
          # Non-Maryland locations
         str_detect(origin, "(?i)(^va\\.?$|virginia|^w\\.? ?va|washington.*d\\.?c|^del|^dc$|florida|^woodbine|^alabama)") ~ "Outside Maryland",
        str_detect(origin, "(?i)Not known") ~ "Unknown",
        str_detect(origin, "(?i)N/A") ~ "Unknown",
        str_detect(origin, "(?i)na") ~ "Unknown",
        str_detect(origin, "(?i)pinna") ~ "Unknown",
        str_detect(origin, "(?i)missing") ~ "Unknown",
       str_detect(origin, "Pinna former residence Jamestown, Md.") ~ "Anne Arundel County",
          is.na(origin) ~ "Unknown",
               TRUE ~ origin
         )) |> 
      mutate(county_origin = str_remove_all(county_origin, ", Md.")) 
```
#st marys? 

```{r}
dc |> 
  count(county_origin) |> 
  arrange(desc(n))


```

Map
```{r}
library(tidycensus)
#subset the geometry
map <- get_acs(geography = "county", 
              variables = c(number_households = "B19001_001", median_income = "B19013_001"), 
               state = "MD", 
               year = 2020, geometry = TRUE) %>% 
  mutate(NAME = str_remove_all(NAME, ", Maryland")) |> 
  select(GEOID, NAME) |> 
  distinct()

#join geocoordinates with baltcity
dc_map <- dc |>  
 inner_join(map, by=c("county_origin"="NAME"))


```
#save map file
```{r}

saveRDS(dc_map, "output/dc_map_nov_26.rds")

```

```{r}
dc_map <- dc_map |> 
  st_as_sf() |> 
  sf::st_transform(4326) |> 
  add_count(county_origin, name = "county_count")
```




# Updated map
```{r}
# Use colorFactor for categorical data
pal <- colorFactor(
  palette = "viridis",  # or "Set1", "Dark2", etc for categories
  domain = dc_map$county_count
)
# Create summary data
summary_data <- dc_map %>%
  count(county_origin) %>%
  filter(county_origin %in% c("Maryland", "Unknown", "Outside Maryland"))

# Create HTML table
table_html <- paste0(
  '<table style="background-color: white; padding: 10px; border: 2px solid gray;">',
  '<tr><th style="text-align: left; padding: 5px;">Origin</th>',
  '<th style="text-align: right; padding: 5px;">Count</th></tr>',
  '<tr><td style="padding: 5px;">Maryland</td><td style="text-align: right; padding: 5px;">45</td></tr>',
  '<tr><td style="padding: 5px;">Unknown</td><td style="text-align: right; padding: 5px;">25</td></tr>',
  '<tr><td style="padding: 5px;">Outside Maryland</td><td style="text-align: right; padding: 5px;">7</td></tr>',
  '</table>'
)

# Create headline
headline_html <- paste0(
  '<div style="background-color: white; padding: 10px; border: 2px solid #333; border-radius: 5px; box-shadow: 0 0 10px rgba(0,0,0,0.3); text-align: center;">',
  '<h3 style="margin: 0; color: #333;">Birth Places of Deceased Boys at Maryland Institutions</h2>',
  '</div>'
)

leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addControl(html = headline_html, position = "topright") %>%  # or "topleft"
  addPolygons(
    color = ~pal(county_count),
    weight = 2.5,
    fillOpacity = 0.5,
    label = ~paste0("Count of deceased boys from ", county_origin, ": ", county_count)
  ) %>%
  addControl(html = table_html, position = "bottomleft")

leafMap
saveWidget(leafMap, "inmates_map_nov_18.html")
```
