---
title: "bridget playtime"
output: html_document
---

# load libraries
```{r echo=FALSE}
library(tidyverse)
library(tidycensus)
library(janitor)
library(dplyr)
```

#load spending data
```{r echo=FALSE}

spending <- read_csv("../data/federal_spending_by_county_2020 2024 rob_0416_v2.csv") |> 
     clean_names()
```



#Make sure each year has the same number of states
```{r echo=FALSE}
options(scipen = 999)  # Disable scientific notation globally

data_2020 <- spending |> filter(fyear==2020)
data_2021 <- spending |> filter(fyear==2021)
data_2022 <- spending |> filter(fyear==2022)
data_2023 <- spending |> filter(fyear==2023)
data_2024 <- spending |> filter(fyear==2024)


nrow(data_2020 |> distinct(fyear, state_nm))
nrow(data_2021 |> distinct(fyear, state_nm))
nrow(data_2022 |> distinct(fyear, state_nm))
nrow(data_2023 |> distinct(fyear, state_nm))
nrow(data_2024 |> distinct(fyear, state_nm))


data_2021 |> filter(!(data_2021$state_nm %in% data_2020$state_nm)) |> distinct(state_nm)
data_2022 |> filter(!(data_2022$state_nm %in% data_2020$state_nm)) |> distinct(state_nm)
```
#Transactions with NA City and NA State by year by agency when that number of transactions is > 1
```{r echo=FALSE}
options(scipen = 999)  # Disable scientific notation globally

spending |>
  filter(is.na(cty_name) & is.na(state_nm)) |>
  group_by(fyear, agency) |>
  summarize (
    Transactions = n()
  ) |> filter(Transactions > 1)

```

#Transactions with NA state but not NA city by year by agency
```{r}
options(scipen = 999)  # Disable scientific notation globally

spending |>
  filter(is.na(state_nm) & !is.na(cty_name)) |>
  group_by(fyear, agency) |>
  summarize (
    cty_name = cty_name,
    Transactions = n()
    
  )


```
#Transactions with NA city but not NA state by year by agency
```{r}
options(scipen = 999)  # Disable scientific notation globally

spending |>
  filter(!is.na(state_nm) & is.na(cty_name)) |>
  group_by(fyear, agency) |>
  summarize (
    Transactions = n()
  )


```

#Transaction number and totals by agency by year where the state is not NA
```{r}
options(scipen = 999)  # Disable scientific notation globally

spending |>
  group_by(fyear, agency) |>
  summarize (
    Transactions = n(),
    Spending_Total = sum(spending_total[!is.na(state_nm)], na.rm = TRUE)
  )
  

```
