---
title: "LCA "
author: "Wells"
date: "2026-05-28"
output: html_document
---

FOR EMILY WILSON
Memo from your newsroom describing the project:

We recently released our 2026 Ad Impressions Study, an end-user/consumer survey that measures the effectiveness/longevity of promotional products, in comparison to other advertising mediums.
 
I’ve attached the raw Ad Impressions data, and here’s the article we published about the study: https://members.asicentral.com/news/strategy/march-2026/asi-s-2026-ad-impressions-study-5-compelling-takeaways/
 
We were thinking it would be cool to come up with a sustainability angle for this data. Perhaps there’s a way to look at the materials/resources required to create a promo product and measure that against the longevity of the product and number of impressions it generates in its lifetime to determine its environmental impact? Maybe she could pick a couple of the top products to examine more closely?

Promo apparel companies with impact calculators on their sites:
https://everywhereapparel.com/pages/sustainability
https://allmade.com/pages/impact



More details from your newsroom are here:
https://docs.google.com/document/d/1Y5me4y6shnXXO5rHXkZjO4r7HTcAU-EZz18jxrTC24Q/edit?usp=sharing

The link to your Google Drive folder
https://drive.google.com/drive/folders/1eAMgj7uF30g1_iggAq2yF4-TMxrbgSju?usp=sharing


Your DJNF GitHub repository
https://github.com/djnf-data-2026/emily-wilson


```{r include=FALSE}
library(tidyverse)
library(readxl)
library(googlesheets4)
library(googledrive)
library(janitor)
```

# Import Ad impressions data
```{r}
#this data is weird - you'll needs to extract specific tables separately
us_ads <- rio::import("https://github.com/djnf-data-2026/emily-wilson/raw/refs/heads/main/data/2026%20Ad%20Impressions%20US%20Products.xlsx") 

```


#Extract specific table
Which of the following types of promotional products would you be most excited to receive?
```{r}
#this process involves so trial and error
#examine the spreadsheet and then adjust the row numbers to extract the table you want:https://docs.google.com/spreadsheets/d/1vcSOjyVHvJNzR8Oe1bj8A9ZOWlKpBFm5/edit?usp=sharing&ouid=102324743793755798467&rtpof=true&sd=true

excited <- us_ads[39:62, ] 
colnames(excited) <- excited[1, ]

excited <- excited[-1, ] |> 
  clean_names() |> 
  as.tibble()

excited <- excited[-1, ] |> 
  mutate(across(2:21, as.numeric)) |> 
  rename(item = na)

glimpse(excited)



```

Import 2026 Ad impressions countries with USA.xlsx
```{r}

```


Import 2026 Ad impressions Canada.xlsx
```{r}

```



# Import LCA data

After some poking around, I think this Bonsai project has the most user friendly LCA data
https://bonsai.uno/gtdr

Use this web interface to check on cotton:
https://www.bonsai.uno/?mode=product&metric=GWP100&version=v2.1.6&product=C_Cotton&region=US&advanced=1&unit=kg

Here's a download of the full dataset
```{r}

footprint <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/emily-wilson/refs/heads/main/data/footprints-v2.1.6.csv?token=GHSAT0AAAAAADWDAQPHELRAA7Q4LC4JQ25Y2QTERTA") |> 
  clean_names()

glimpse(footprint)

```
## Dataset description
From Claude:

1,538 products/sectors × 54 countries × 2 metrics
Metric	Full Name	What it measures
GWP100	Global Warming Potential, 100-year	Standard carbon accounting — cumulative heat trapped over 100 years
GTP20	Global Temperature Potential, 20-year	Temperature change at 20 years — weights short-lived gases (like methane) much more heavily
GWP100 is the standard used in most climate policy (Paris Agreement, IPCC reports). GTP20 is more useful if you're interested in near-term climate impacts.

Your data uses C_ prefix (commodity) while this page shows M_ (market for).

Filter to one metric for most analyses:

footprint_gwp <- footprint |> filter(metric == "GWP100")
footprint_gtp <- footprint |> filter(metric == "GTP20")

An explanation of the data fields
https://search.lca.aau.dk/web/concept/https%3A//glossary.bonsai.uno/Bonsut-v2.1.3-flowobjects/fm?concept_scheme=https%3A%2F%2Fglossary.bonsai.uno%2FBonsut-v2.1.3-flowobjects&language=en

Exiobase defines the variables
https://zenodo.org/records/14614930
--EXIOBASE 3: For best in class environmental-economic accounting data. Get insight into global supply-chains and the environmental impacts of consumption.

More on Exiobase
https://epub.wupperinst.org/frontdoor/deliver/index/docId/6938/file/6938_Stadler.pdf


# Pokin around

```{r}
footprint |> 
  filter(metric == "GWP100") |> 
  filter(str_detect(flow_code, "C_Cotton"))

#write_csv(precincts, "../data/precincts.csv")
```

Claude explanation
1. It's:
"€1 million worth of cotton production in Australia generates -3,096.6 tonnes CO2eq"

The value is always tonnes CO2eq per million euros of output — emissions are always on the left side of that ratio.

2. The negative value

In input-output analysis, negative = net carbon sequestration. For cotton it makes sense because:

Cotton plants absorb CO2 from the atmosphere as they grow
If the carbon stored in the fiber exceeds the emissions from farming, transport, processing etc., the net footprint is negative
The model is saying Australian cotton production removes more carbon than it emits
This is called a carbon sink and shows up in agriculture and forestry products fairly often in these models.



### Percent of the data that is negative
```{r}
footprint |>
  filter(metric == "GWP100") |>
  summarise(
    pct_negative = mean(value < 0) * 100,
    n_negative   = sum(value < 0)
  )


```

