---
title: "LCA "
author: "Wells"
date: "2026-05-19"
output: html_document
---

FOR EMILY WILSON
Memo from your newsroom describing the project:

We recently released our 2026 Ad Impressions Study, an end-user/consumer survey that measures the effectiveness/longevity of promotional products, in comparison to other advertising mediums.
 
I’ve attached the raw Ad Impressions data, and here’s the article we published about the study: https://members.asicentral.com/news/strategy/march-2026/asi-s-2026-ad-impressions-study-5-compelling-takeaways/
 
We were thinking it would be cool to come up with a sustainability angle for this data. Perhaps there’s a way to look at the materials/resources required to create a promo product and measure that against the longevity of the product and number of impressions it generates in its lifetime to determine its environmental impact? Maybe she could pick a couple of the top products to examine more closely?

More details from your newsroom are here:
https://docs.google.com/document/d/1Y5me4y6shnXXO5rHXkZjO4r7HTcAU-EZz18jxrTC24Q/edit?usp=sharing


```{r}
library(tidyverse)
library(readxl)
library(googlesheets4)
library(googledrive)
library(janitor)
```



#Import data

After some poking around, I think this Bonsai project has the most user friendly LCA data
https://bonsai.uno/gtdr


```{r}

footprint <- read_csv("/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Emily Wilson/footprints-v2.1.6.csv") |> 
  clean_names()

glimpse(footprint)

```

So the dataset is:

1,538 products/sectors × 54 countries × 2 metrics = theoretically 166,104 combinations — close to your 188,924 rows, so coverage is fairly complete but not every product exists in every country
GTP20 is worth knowing about:

Metric	Full Name	What it measures
GWP100	Global Warming Potential, 100-year	Standard carbon accounting — cumulative heat trapped over 100 years
GTP20	Global Temperature Potential, 20-year	Temperature change at 20 years — weights short-lived gases (like methane) much more heavily
GWP100 is the standard used in most climate policy (Paris Agreement, IPCC reports). GTP20 is more useful if you're interested in near-term climate impacts.

You'll likely want to filter to one metric for most analyses:

footprint_gwp <- footprint |> filter(metric == "GWP100")
footprint_gtp <- footprint |> filter(metric == "GTP20")


Use this web interface to check on cotton:
https://www.bonsai.uno/?mode=product&metric=GWP100&version=v2.1.6&product=C_Cotton&region=US&advanced=1&unit=kg



An explanation of the data fields
https://search.lca.aau.dk/web/concept/https%3A//glossary.bonsai.uno/Bonsut-v2.1.3-flowobjects/fm?concept_scheme=https%3A%2F%2Fglossary.bonsai.uno%2FBonsut-v2.1.3-flowobjects&language=en


# Pokin around

```{r}
footprint |> 
  filter(metric == "GWP100") |> 
  filter(str_detect(flow_code, "Cotton"))

#write_csv(precincts, "../data/precincts.csv")
```


```{r}

library(rvest)

page <- read_html("https://search.lca.aau.dk/web/concept/https%3A//glossary.bonsai.uno/Bonsut-v2.1.3-flowobjects/fm?concept_scheme=https%3A%2F%2Fglossary.bonsai.uno%2FBonsut-v2.1.3-flowobjects&language=en")

# Extract table if present
flow_labels <- page |> html_table() |> pluck(1)



```


```{r}
top <- primary |> 
  group_by(precinct) |> 
  count(
    dem_president_of_the_united_states_6_500001630106, 
    rep_president_of_the_united_states_6_500001630206
  ) |> 
  filter(
    dem_president_of_the_united_states_6_500001630106 == "Joe Biden (10066066.0000254)" | 
    rep_president_of_the_united_states_6_500001630206 == "Donald J. Trump (10066066.0000246)"
  ) |> 
  pivot_longer(
    cols = c(
      dem_president_of_the_united_states_6_500001630106, 
      rep_president_of_the_united_states_6_500001630206
    ),
    names_to = "candidate_type",  # Changed from "precinct" to avoid clashing
    values_to = "candidate_name", # Moves the actual name strings here
    values_drop_na = TRUE         # Optional: Drops the messy 'NA' rows seen in your output
  )
  



```


Wider
```{r}

top_wide <- primary |> 
  group_by(precinct) |> 
  count(
    dem_president_of_the_united_states_6_500001630106, 
    rep_president_of_the_united_states_6_500001630206
  ) |> 
  filter(
    dem_president_of_the_united_states_6_500001630106 == "Joe Biden (10066066.0000254)" | 
    rep_president_of_the_united_states_6_500001630206 == "Donald J. Trump (10066066.0000246)"
  ) |> 
  mutate(candidate = coalesce(
    dem_president_of_the_united_states_6_500001630106, 
    rep_president_of_the_united_states_6_500001630206
  )) |> 
  # 2. Pivot wider using 'id_cols' to force exactly one row per precinct
  pivot_wider(
    id_cols = precinct,
    names_from = candidate,
    values_from = n
  ) |> 
  rename(Biden = 2, Trump =3)

```


Chart
```{r}
top_wide |> 
  pivot_longer(
    cols = -precinct, 
    names_to = "candidate", 
    values_to = "votes"
  ) |> 
  mutate(candidate = case_when(
    stringr::str_detect(candidate, "Biden") ~ "Joe Biden",
    stringr::str_detect(candidate, "Trump") ~ "Donald Trump",
    TRUE ~ candidate
  )) |> 
  ggplot(aes(x = precinct, y = votes, fill = candidate)) +
  geom_col(position = "dodge") +
  labs(
    title = "Primary candidate votes",
    caption = "Graphic by Rob Wells, 5-20-2026",
    y = "Votes",
    x = "Precinct",
    fill = "Candidate"
  ) +
    theme(axis.text.x = element_text(angle=90, size=6)) 
```


#Notes only - 

# Wrangling to get it into R

#GENERAL_11PG24_Prov_CVR_SBE - 1
```{r}
general1 <- read_excel("/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Katelynn Winebrenner/2024 Cast Vote Record/GENERAL_11PG24_Prov_CVR_SBE - 1.xlsx")

write_csv(general1, "../data/GENERAL_11PG24_Prov_CVR_SBE_1.csv")
```

GENERAL_11PG24_Prov_CVR_SBE
```{r}
general2 <- read_excel("/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Katelynn Winebrenner/2024 Cast Vote Record/GENERAL_11PG24_Prov_CVR_SBE.xlsx")


write_csv(general2, "../data/GENERAL_11PG24_Prov_CVR_SBE.csv")
```

PRIMARY_11PP24_Prov_CVR_SBE
```{r}

primary <- read_excel("/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Katelynn Winebrenner/2024 Cast Vote Record/PRIMARY_11PP24_Prov_CVR_SBE.xlsx")


write_csv(primary, "../data/PRIMARY_11PP24_Prov_CVR_SBE.csv")
```
