---
title: "Government Spending Master DJNF"
author: "Wells"
date: "2026-05-15"
output: html_document
---

# load libraries
```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
```

#load spending data
```{r}

spending <- read_csv("../data/federal_spending_by_county_2020 2024 rob_0416_v2.csv") |> 
     clean_names()

```

**pokin around!**

## Spending by agency

```{r}
options(scipen = 999)  # Disable scientific notation globally

spending |> 
group_by(agency) |> 
  summarize(
    n = n(),
    total_spent = sum(spending_total, na.rm = TRUE)
  ) |> 
  arrange(desc(total_spent))

```

## Spending by state

```{r}
spending |> 
group_by(state_nm) |> 
  summarize(
    n = n(),
    total_spent = sum(spending_total, na.rm = TRUE)
  ) |> 
  arrange(desc(total_spent))

```

## Spending by city
```{r}
spending |> 
group_by(cty_name, state_nm) |> 
  summarize(
     total_spent = sum(spending_total, na.rm = TRUE),
     n = n()
  ) |> 
  arrange(desc(total_spent))

```

## Highlight specific city
```{r}

spending |> 
  filter(cty_name == "HENNEPIN") |> 
  group_by(agency) |> 
  summarize(
     total_spent = sum(spending_total, na.rm = TRUE),
     n = n()
  ) |> 
  arrange(desc(total_spent))
```

## Highlight specific city and agency
```{r}
spending |> 
  filter(cty_name == "HENNEPIN") |> 
  filter(agency == "Department of Health and Human Services") |> 
  select(fyear, cty_name, spending_total)
```

To drill down into specific spending by county, we now go to USASpending.gov: https://www.usaspending.gov/
* Select start searching awards
* Select one fiscal year (multiple years = too much data)
* Select location, place of performance and type in the county name
* Search!

The web page serves up a form with spending by specific company in that county per year. Gold mine!
Make sure this is the search you want. 
To download this data, see the download icon in the black toolbar


Law enforcement /defense  spending by state. 
Examine the trend from 2020-2024 for your state. Has that changed over time? How does your state compare to others?
Using the USAspending.gov website, name the top five companies by this category by state
Which counties are the winners? Understand there are flaws with county-level data


```{r}



```

Social services spending by county. 
Examine the trend from 2020-2024 for your state. Has that changed over time? How does your state compare to others?
Using the USAspending.gov website, name the top five companies by this category by state
Which counties are the winners? Understand there are flaws with county-level data
Compare to law enforcement

```{r}



```


Differences across administrations
For your state, compare the last year of Trump’s first administration (2020) to the last year of the Biden administration (2021)
How did spending differ overall? Was one administration spending way more/less than another?
Was one administration pouring more/less into specific agencies than the other administration?

```{r}



```

