---
title: "Laine-Hospital Data"
author: "Wells"
date: "2026-05-31"
output: html_document
---


Examine hospital expenditures over time by county, the effects of consolidation

# Load packages - do this once

```{r}
# install.packages("tidyverse")
# install.packages("janitor")
# install.packages("rio")
# install.packages("googledrive")

```


```{r}
library(tidyverse)
library(janitor)
library(rio)
library(googledrive)


```

# Loading data AHRFSN data for 2024 and 2025
**need documentation files**
```{r}
ahrfsn2025 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/ahrfsn2025.csv")

ahrfsn2024 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/ahrfsn2024.csv")

glimpse(ahrfsn2025)

```


# Loading AHRF 2025 and 2024 files for Michigan only

```{r}

AHRF2025_A <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/ahrfsn2024.csv")


AHRF2024_A <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/ahrfsn2024.csv")
```



Load healthcare market saturation csv 

Some of the HSRA hospital data is saved as ACS files, unfortunately. I have the documentation but I'm going to have to edit it so I can easily add the column names as it is not in a nicely joinable format. I might need some help with that during the training

I tried for quite some time to save each HSRA file properly through Visual Studio Code and R, but it's going to take longer than I thought given the scale and the documentation issues. Perhaps I'd need to heavily cut the HRSA data before looking into it because having every single county in the U.S. for all of those rows is jamming it up quite a bit.


# Load AHRF2025 file
```{r}
googledrive::drive_auth()
drive_download(
  as_id("1K8DzTY5yTNI0EeIUpadeCrE552tcxRsj"),  # from the share link
  path = "../data/AHRF2025.csv",
  overwrite = TRUE
)
AHRF2025 <- read_csv("../data/AHRF2025.csv")|> 
  clean_names()

AHRF2025_MI <- AHRF2025 |> 
  filter(st_name_abbrev == "MI")

#write_csv(AHRF2025_MI, "../data/AHRF2025_MI.csv" )

```

# Load AHRF2024 file
```{r}
#https://drive.google.com/file/d/1GgVgzo7DSv1pn8BjCuhygmILEjGT04Ja/view?usp=sharing
googledrive::drive_auth()
drive_download(
  as_id("1GgVgzo7DSv1pn8BjCuhygmILEjGT04Ja"),  # from the share link
  path = "../data/AHRF2024.csv",
  overwrite = TRUE
)
AHRF2024 <- read_csv("../data/AHRF2024.csv")|> 
  clean_names()

AHRF2024_MI <- AHRF2024 |> 
  filter(st_name_abbrev == "MI")

write_csv(AHRF2024_MI, "../data/AHRF2024_MI.csv" )

```


What are .ASC & .ASCII files?
An ASC file is a variant of the ASCII format, which is an encryption file used by Pretty Good Privacy (PGP) for secure online communication. It includes messages and digital signatures, as well as plain text and binary information. It also includes a verification key. It’s mainly used to post online security notices and to transmit messages by text and email.