---
title: "Voting-Frederick"
author: "Wells"
date: "2026-05-27"
output: html_document
---
# Frederick County voting analysis

You have four folders of Frederick County election data to conduct an analysis of the outcomes of the county council, school board and county executive races: 2018 results, 2022 cast vote record; 2022 results; 2024 cast vote records.; council district and election boundaries.

We are starting with the 2024 results for general election and primary election.
The plan is to get a workflow down to produce a map and some charts that describe top Republican and Democratic precincts for the county council, school board and county executive races.
And see if there was any change from the primary to the general election and examine the disparity in voting by political party for national versus local offices.
Then compare those margins of victory to the presidential and senate races. 

Hopefully, you will be able to compare these 2024 results to the 2022 races, but that comparison will be problematic since non-presidential years lack the same turnout


```{r}
library(tidyverse)
library(readxl)
library(googlesheets4)
library(googledrive)
library(janitor)
```
#Import data
```{r}

general1 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/GENERAL_11PG24_Prov_CVR_SBE_1.csv") |> 
  clean_names()


general2 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/GENERAL_11PG24_Prov_CVR_SBE.csv") |> 
  clean_names()

primary <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/PRIMARY_11PP24_Prov_CVR_SBE.csv") |> 
  clean_names()

```

# Pokin around

```{r}
precincts <- primary |> 
  group_by(precinct) |> 
  count()

#write_csv(precincts, "../data/precincts.csv")
```



```{r}
top <- primary |> 
  group_by(precinct) |> 
  count(
    dem_president_of_the_united_states_6_500001630106, 
    rep_president_of_the_united_states_6_500001630206
  ) |> 
  filter(
    dem_president_of_the_united_states_6_500001630106 == "Joe Biden (10066066.0000254)" | 
    rep_president_of_the_united_states_6_500001630206 == "Donald J. Trump (10066066.0000246)"
  ) |> 
  pivot_longer(
    cols = c(
      dem_president_of_the_united_states_6_500001630106, 
      rep_president_of_the_united_states_6_500001630206
    ),
    names_to = "candidate_type",  # Changed from "precinct" to avoid clashing
    values_to = "candidate_name", # Moves the actual name strings here
    values_drop_na = TRUE         # Optional: Drops the messy 'NA' rows seen in your output
  )
  



```


Wider
```{r}

top_wide <- primary |> 
  group_by(precinct) |> 
  count(
    dem_president_of_the_united_states_6_500001630106, 
    rep_president_of_the_united_states_6_500001630206
  ) |> 
  filter(
    dem_president_of_the_united_states_6_500001630106 == "Joe Biden (10066066.0000254)" | 
    rep_president_of_the_united_states_6_500001630206 == "Donald J. Trump (10066066.0000246)"
  ) |> 
  mutate(candidate = coalesce(
    dem_president_of_the_united_states_6_500001630106, 
    rep_president_of_the_united_states_6_500001630206
  )) |> 
  # 2. Pivot wider using 'id_cols' to force exactly one row per precinct
  pivot_wider(
    id_cols = precinct,
    names_from = candidate,
    values_from = n
  ) |> 
  rename(Biden = 2, Trump =3)

```


Chart
```{r}
top_wide |> 
  pivot_longer(
    cols = -precinct, 
    names_to = "candidate", 
    values_to = "votes"
  ) |> 
  mutate(candidate = case_when(
    stringr::str_detect(candidate, "Biden") ~ "Joe Biden",
    stringr::str_detect(candidate, "Trump") ~ "Donald Trump",
    TRUE ~ candidate
  )) |> 
  ggplot(aes(x = precinct, y = votes, fill = candidate)) +
  geom_col(position = "dodge") +
  labs(
    title = "Primary candidate votes",
    caption = "Graphic by Rob Wells, 5-20-2026",
    y = "Votes",
    x = "Precinct",
    fill = "Candidate"
  ) +
    theme(axis.text.x = element_text(angle=90, size=6)) 
```


#Notes only - 

# Wrangling to get it into R

#GENERAL_11PG24_Prov_CVR_SBE - 1
```{r}
general1 <- read_excel("/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Katelynn Winebrenner/2024 Cast Vote Record/GENERAL_11PG24_Prov_CVR_SBE - 1.xlsx")

write_csv(general1, "../data/GENERAL_11PG24_Prov_CVR_SBE_1.csv")
```

GENERAL_11PG24_Prov_CVR_SBE
```{r}
general2 <- read_excel("/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Katelynn Winebrenner/2024 Cast Vote Record/GENERAL_11PG24_Prov_CVR_SBE.xlsx")


write_csv(general2, "../data/GENERAL_11PG24_Prov_CVR_SBE.csv")
```

PRIMARY_11PP24_Prov_CVR_SBE
```{r}

primary <- read_excel("/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Katelynn Winebrenner/2024 Cast Vote Record/PRIMARY_11PP24_Prov_CVR_SBE.xlsx")


write_csv(primary, "../data/PRIMARY_11PP24_Prov_CVR_SBE.csv")
```
