---
title: "Government Spending Master DJNF"
author: "Wells"
date: "2026-05-01"
output: html_document
---

# load libraries
```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
```

#load spending data
```{r}

spending <- read_csv("../data/federal_spending_by_county_2020 2024 rob_0416_v2.csv") |> 
     clean_names()

```

**pokin around!**

## Spending by agency

```{r}
options(scipen = 999)  # Disable scientific notation globally

spending |> 
group_by(agency) |> 
  summarize(
    n = n(),
    total_spent = sum(spending_total, na.rm = TRUE)
  ) |> 
  arrange(desc(total_spent))

```

## Spending by state

```{r}
spending |> 
group_by(state_nm) |> 
  summarize(
    n = n(),
    total_spent = sum(spending_total, na.rm = TRUE)
  ) |> 
  arrange(desc(total_spent))

```

## Spending by county
```{r}
spending |> 
group_by(cty_name, state_nm) |> 
  summarize(
     total_spent = sum(spending_total, na.rm = TRUE),
     n = n()
  ) |> 
  arrange(desc(total_spent))

```

## Highlight specific county
```{r}

spending |> 
  filter(cty_name == "HENNEPIN") |> 
  group_by(agency) |> 
  summarize(
     total_spent = sum(spending_total, na.rm = TRUE),
     n = n()
  ) |> 
  arrange(desc(total_spent))
```

## Highlight specific county and agency
```{r}
spending |> 
  filter(cty_name == "HENNEPIN") |> 
  filter(agency == "Department of Health and Human Services") |> 
  select(fyear, cty_name, spending_total)
```



Who are these employers
https://www.forbes.com/lists/best-employers-by-state/

zdiaz@umd.edu Zaida Diaz
Hi Zaida - I saw your name on a business research guide. I hope you are doing well.
I am leading a group of journalism students in June to examine federal spending by state and county. They will analyze data and come up with things like Hennepin Minnesota leads the list of cities with the most federal grants from 2020-2024. Hennepin? It's the home of the Mayo Clinic, which raked in billions of federal money from HHS.

So with that in mind, I am looking for a handy lookup database to determine largest employers by city, county or state, and it would have to include private businesses and nonprofits (the SEC's Edgar system wouldn't cut it, for example). ValueLine and Morningstar are not useful for this. Maybe it's buried in some corner of Factiva?

So your help would be most welcome! Thank you, Rob Wells


Hi Liu - I hope you're doing well. Thanks again for sending along this dataset. Do you have time next week for a few followup questions? This data appears to be quite clean and really useful for this project. 

But I am not understanding what is going on here. I ran a few queries and got some very weird results. 

Hennepin County (Minneapolis) in Minnesota had $706 billion in total federal spending over the five-year period, double that of Los Angeles County, even though LA has about 8 times the population. 
The Hennepin spending was basically from HHS, but any idea who actually got that money? I would guess UnitedHealth Care, but most of its money comes from the VA.  
I'm trying to figure out that part of the puzzle -- I assume you extracted the data from the https://www.usaspending.gov/ ?
Thanks so much, Rob

